---
title: "BIS015L Group Project On UFO Sightings: Data Processing, Analysis, and Visualization Script"
output: 
  html_document: 
    keep_md: true
date: "2024-03-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installing & loading any required packages:
```{r}
#install.packages("paletteer")
```

```{r}
library("tidyverse")
library("naniar")
library("janitor")
library("ggmap")
library("dplyr")
```

## Pre-processing the data set:

### Importing the data and assigning "NA" to any missing values
```{r}
ufo <- readr::read_csv("ufo_data/complete.csv", na = c("", "unkown")) %>% clean_names
```

### Inspecting its basic structure
```{r}
glimpse(ufo)
```

```{r}
summary(ufo)
```

### Removing unnecessary data, seperating dates, changing data types, and filtering out anything not in the US
```{r}
ufo2 <- ufo %>%
  separate(datetime, into = c("month", "day","year"), sep="/") %>%
  separate(year, into = c("year", "time"), sep = " ") %>%
  select(-duration_hours_min, -comments, -date_posted) %>%
  filter(country=="us", latitude!=0, longitude!=0,  !is.na(latitude), !is.na(longitude)) %>%
  mutate(year=as.factor(year)) %>%
  mutate(latitude=as.numeric(latitude)) %>%
  mutate(longitude=as.numeric(longitude)) %>%
  mutate(shape=as.factor(shape)) %>%
  mutate(state=as.factor(state))
```
## Creating a map of UFO sightings in the US
### Loading API key
```{r}
register_stadiamaps("aff428bd-8ad7-4b69-8b6b-b34301483ba9", write = FALSE)
```

### Creating a bounding box for the map
```{r}
ufo2 %>% 
  select(latitude, longitude) %>% 
  summary()
```

```{r}
latitude <- c(17.97, 70.67)
longitude <- c(-176.66, -65.83)
alien_bbox <- make_bbox(longitude, latitude, f = 0.03)
```

### Load the map from Stamen
```{r}
alien_map <-  get_stadiamap(alien_bbox, maptype = "stamen_terrain", zoom=5)
```
```{r}
ggmap(alien_map)
```

### Compiling the final map
```{r}
ggmap(alien_map) + 
  geom_point(data = ufo2, aes(longitude, latitude), size=0.1) +
  labs(x= "Longitude", y= "Latitude", title="UFO Sightings in the U.S (1910 - 2014)")
```
## Ploting the number of UFO sightings over the years
```{r}
str(ufo2)
```

### Creating a summary table of the yearly sightings listed in descending order
```{r}
ufo2 %>% 
  group_by(year) %>% 
  summarise(sightings = n()) %>% 
  arrange(desc(sightings))
```

```{r}
sightings_summary <- ufo2 %>%
  filter(!is.na(year)) %>% 
  group_by(year) %>% 
  summarise(sightings = n())
```

### Compiling the plot
```{r}
sightings_summary$sightings <- as.numeric(sightings_summary$sightings)
```


```{r}
ggplot(sightings_summary, aes(x = year, y = sightings, group = 1))+
  geom_point(size = 1)+
  geom_path(color = "blue")+
  labs(title = "UFO Sightings Over the Years (1910-2014)",
       x= "Year",
       y = "Number of Sightings")+
  scale_x_discrete(breaks = seq(1910,2014, by = 3))+
  theme_minimal()+
  theme(axis.text.x=element_text(size= 7.5,angle=70, hjust=1))
```

## Plotting the distribution of sighting duration in each state
### Creating a summary table of the average sighting duration in each state
```{r}
ufo2_avg_time <- ufo2 %>% 
  filter(!is.na(duration_seconds & duration_seconds != 0) & !is.na(shape)) %>% 
  group_by(state, shape) %>% 
  summarise(avg_secs = mean(duration_seconds)) %>%
  arrange(desc(avg_secs))
ufo2_avg_time
```

### Compiling the plot
```{r}
ufo2_avg_time %>%
  ggplot(aes(x = state, y = avg_secs))+
  geom_col()+
  labs(title = "Average Duration of UFO Sightings by State",
       x = "State",
       y= "Duration (Seconds)")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
ggplot(ufo2, aes(x = longitude, y = latitude)) +
  geom_tile(stat = "density2d", aes(fill = ..density..), color = "white") +
  scale_fill_viridis_c() +  # Choose a color scale
  labs(title = "Heatmap of UFO Sightings", x = "Longitude", y = "Latitude", fill = "Density") +
  theme_minimal()
```

## Plotting the total number of sightings in each state
### Creating a summary table of the total recorded sightings in each state
```{r}
ufo2 %>% 
  group_by(state) %>% 
  summarise(n_sightings = n()) %>%
  arrange(desc(n_sightings))
```

### Compiling the plot
```{r}
ufo2 %>% 
  group_by(state) %>% 
  summarise(n_sightings = n()) %>%
  arrange(desc(n_sightings)) %>%
  ggplot(aes(x = state, y = n_sightings))+
  geom_col()+
  labs(title = "Total Recorded UFO Sightings by State",
       x = "State",
       y= "Number of Sightings")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Plotting the frequency of sightings organized by shape
### Creating a summary table of the total recorded sightings in each state
```{r}
ufo2 %>% 
  group_by(shape) %>% 
  summarise(n_sightings = n()) %>%
  arrange(desc(n_sightings))
```

### Compiling the plot
```{r}
ufo2 %>% 
  filter(!is.na(shape)) %>%
  group_by(shape) %>% 
  summarise(n_sightings = n()) %>%
  arrange(desc(n_sightings)) %>%
  ggplot(aes(x = shape, y = n_sightings))+
    geom_col()+
    labs(title = "Total Recorded UFO Sightings by Sighting Shape",
       x = "Shape",
       y= "Number of Sightings")+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
#why are there still unknowns here?
```

## ShinyApp

```{r}
#install.packages("shinythemes")
library(shinythemes)
```

```{r}
library(shiny)
library(shinydashboard)
```
```{r}
ui <- fluidPage(
  theme = shinytheme("yeti"),
  # App title 
  tags$head(
    tags$style(
      HTML("
           .title {
             font-family: 'Chewy', cursive;
             font-size: 24px;
           }
           "
      )
    ),
  ),
  headerPanel(
    div(class = "title", "UFO Tracker")
  ),

  # Sidebar panel for inputs
  sidebarPanel(
    selectInput("shape", "Select UFO Shape", choices = c(unique(ufo2$shape))),
    selectInput("state", "Select State", choices = c(unique(ufo2$state)))
  ),
  
  # Main panel for displaying outputs 
  mainPanel(
    plotOutput("plot")
  )
)

# Define server logic 
server <- function(input, output) {
  
  # Render plot based on selected UFO shape and state
  output$plot <- renderPlot({
    req(input$shape, input$state)  # Require both inputs to be selected
    
    filtered_data <- ufo2 %>%
      filter(shape == input$shape & state == input$state)
    
    if (nrow(filtered_data) == 0) {
      # If no data found, display a message
      plot(1, type = "n", axes = FALSE, xlab = "", ylab = "", main = "No data available")
    } else {
      # Otherwise, plot the data
      ggplot(filtered_data, aes(x = month)) +
        geom_bar(fill = "skyblue") +
        labs(title = paste("UFO Sightings for Shape:", input$shape, "and State:", input$state),
             x = "Month",
             y = "Number of Sightings") +
        theme_linedraw()
    }
  })
}

# Start the app
shinyApp(ui, server)
```



